---
title: "Redin Ireland et al. pseudotime improved"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output:
  html_document:
    highlight: tango
    df_print: paged
    toc: yes
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    code_folding: hide
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, result='hold',tidy=TRUE, cache=FALSE, error = T, echo = F)
knitr::opts_knit$set(progress=F,verbose=FALSE)

```

```{r}
#Libraries needed for the analysis
require(dplyr)
require(base)
require(Seurat)
require(patchwork)
require(stringr)
require(plyr)
require(data.table)
require(ggplot2)
require(Matrix)
library(knitr)
library(UCell)
library(clusterProfiler)
library(msigdbr)
library(monocle)
library(Seurat)
library(fgsea)
library(tidyr)
```


```{r}
#load both objects from https://duke.app.box.com/s/xof67knrdpf21q89tysj674tddxfsyke
pseudotime_object <- readRDS("/local/storage/maa7095/Raw_data_projects/scRNAseq_projects/Redin_SCLC/Ireland_Fig4D_RPM_Tumors_Timepoints_OrderedMonocle.rds")
```

```{r}
genes_to_plot <- c("Myc","Mycl","Smarca2","Smarca4", "Yap1","Ascl1","Neurod1","Nrg1","Trp63","Egfr","Erbb3","Erbb2","Sox2","Insm1","B2m","Foxa2","Myt1","Nrsn1","Rtn1","Srrm4","Rest")
#Krt5 doenst show expression in cell lines.
```


Pseudotime is an abstract unit of progress: it's simply the distance between a cell and the start of the trajectory, measured along the shortest path. The trajectory's total length is defined in terms of the total amount of transcriptional change that a cell undergoes as it moves from the starting state to the end state.

# Expression profile in pseudotime coordinates for time points 4,7,11,14, 17 and 21 days with 0s removed. {.tabset}

```{r}
time_interest <- c(4,7,11,14, 17,21)
#colored expression by log10(value + 0.1)


expression_cells_nozeros <- function(object, gene){
  plot_interest <-  plot_cell_trajectory(object, markers = gene,  use_color_gradient = T) + facet_wrap(~time, nrow = 1)
plot_interest$data <- plot_interest$data %>% filter(time %in% time_interest)
plot_interest$data <- plot_interest$data %>% filter(value > 0)
pdf(file = paste0("/SSD/maa7095/scRNAseq/Resin_Esther_sc_small_cell_lung_cancer/SCLC_sc_analysis/Ireland_analysis/plots_pdf/new_script/Exp_cell_lines_0s_removed/",gene, ".pdf") ,   # The directory you want to save the file in
    width = 10, # The width of the plot in inches
    height = 10) # The height of the plot in inches
plot2 <- plot_interest +  scale_colour_gradient(low = "black", high = "green") +labs(color = "Expression") + theme(legend.key.size = unit(1, "cm"))

print(plot2)

dev.off()
} 
```

```{r}
for (gene in genes_to_plot) {
  expression_cells_nozeros(pseudotime_object, gene)
}
```


# {-}


# Expression profile in pseudotime coordinates for time points 4,7,11,14, 17 and 21 days with 0s {.tabset}

```{r}
time_interest <- c(4,7,11,14, 17,21)
#colored expression by log10(value + 0.1)
expression_cells <- function(object, gene){
  plot_interest <-  plot_cell_trajectory(object, markers = gene,  use_color_gradient = T) + facet_wrap(~time, nrow = 1)
plot_interest$data <- plot_interest$data %>% filter(time %in% time_interest)

subset_df <- plot_interest$data %>% filter(value>0)
  subset_df$new_val <- log10(subset_df$value + 0.1)
  pdf(file = paste0("/SSD/maa7095/scRNAseq/Resin_Esther_sc_small_cell_lung_cancer/SCLC_sc_analysis/Ireland_analysis/plots_pdf/new_script/Exp_cell_lines/",gene, ".pdf") ,  # The directory you want to save the file in
    width = 10, # The width of the plot in inches
    height = 10) # The height of the plot in inches
  plot2 <- plot_interest + 
    geom_point(data = subset_df, aes(x = data_dim_1, y = data_dim_2, colour = new_val)) + 
    scale_colour_gradient(low = "black", high = "green") + 
    labs(color = "Expression") + 
    theme(legend.key.size = unit(1, "cm"))
  
  print(plot2)
  dev.off()
} 
```

```{r}
for (gene in genes_to_plot) {
  expression_cells(pseudotime_object, gene)
}
```

# {-}

# NE profile in pseudotime coordinates for time points 4,7,11,14, 17 and 21 days with 0s removed {.tabset}

```{r}
time_interest <- c(4,7,11,14, 17,21)
#colored expression by log10(value + 0.1)
NE_pseudotime_gene <- function(object, gene){
  plot_interest <-  plot_cell_trajectory(object, markers = gene,  use_color_gradient = F, color_by = "NE_spearman",) + facet_wrap(~time, nrow = 1)
  plot_interest$data <- plot_interest$data %>% filter(time %in% time_interest)
  plot_interest$data <- plot_interest$data %>% filter(value > 0)
  plot_interest$data <-  plot_interest$data %>% drop_na(NE_spearman)
  plot_interest$mapping$size <- NULL
  pdf(file = paste0("/SSD/maa7095/scRNAseq/Resin_Esther_sc_small_cell_lung_cancer/SCLC_sc_analysis/Ireland_analysis/plots_pdf/new_script/NE_cell_lines_0s_removed/",gene, ".pdf") ,  # The directory you want to save the file in
    width = 10, # The width of the plot in inches
    height = 10) # The height of the plot in inches
  plot2 <- plot_interest +  scale_colour_gradient(low = "black", high = "brown3") +labs(color = "NE score")  +ggtitle(paste0(gene))  + theme(legend.key.size = unit(1, "cm"))
  print(plot2)
  dev.off()
}
```

```{r}
for (gene in genes_to_plot) {
  NE_pseudotime_gene(pseudotime_object, gene)
}
```


## {-}

## Smarca4 all time points together with 0s removed.

```{r, eval=F}
time_interest <- c(4,7,11,14, 17,21)
#colored expression by log10(value + 0.1)
plot_smarca4_ne <-  plot_cell_trajectory(pseudotime_object, markers = "Smarca4",  use_color_gradient = F, color_by = "NE_spearman")
  plot_smarca4_ne$data <- plot_smarca4_ne$data %>% filter(time %in% time_interest)
  plot_smarca4_ne$data <- plot_smarca4_ne$data %>% filter(value > 0)
  plot_smarca4_ne$data <-  plot_smarca4_ne$data %>% drop_na(NE_spearman)
  plot_smarca4_ne$mapping$size <- NULL
  plot_smarca4_ne +  scale_colour_gradient(low = "black", high = "brown3") +labs(color = "NE score")  +ggtitle("Smarca4") + theme(legend.key.size = unit(1, "cm"))

```
# {-}

```{r}
genes_to_plot <- c("Myc","Mycl","Smarca2","Smarca4", "Yap1","Ascl1","Neurod1","Nrg1","Egfr","Erbb3","Erbb2","Sox2","Insm1","B2m","Foxa2","Myt1","Nrsn1","Rtn1","Srrm4","Rest")
#no expression of Trp63 and Krt5 in tumors
```



# Expression profile in pseudotime coordinates for time points RPM1, RPM2, RPM3 and RPM4 with 0s removed. {.tabset}

```{r}
time_interest <- c("RPM1","RPM2", "RPM3", "RPM4")
#we plot here log10(value + 0.1)
expression_tumors_nozero <- function(object, gene){
  plot_interest <-  plot_cell_trajectory(object, markers = gene,  use_color_gradient = T) + facet_wrap(~time, nrow = 1)
plot_interest$data <- plot_interest$data %>% filter(time %in% time_interest)
plot_interest$data <- plot_interest$data %>% filter(value > 0)
pdf(file = paste0("/SSD/maa7095/scRNAseq/Resin_Esther_sc_small_cell_lung_cancer/SCLC_sc_analysis/Ireland_analysis/plots_pdf/new_script/Expr_tumors_0s_removed/",gene, ".pdf") ,  # The directory you want to save the file in
    width = 10, # The width of the plot in inches
    height = 10) # The height of the plot in inches
plot2 <- plot_interest +  scale_colour_gradient(low = "black", high = "green") +labs(color = "Expression")  +ggtitle(paste0(gene))+ theme(legend.key.size = unit(1, "cm"))
print(plot2)
dev.off()
}
```

```{r}
for (gene in genes_to_plot) {
  expression_tumors_nozero(pseudotime_object, gene)
}
```

# {-}


# Expression profile in pseudotime coordinates for time points RPM1, RPM2, RPM3 and RPM4 with 0s. {.tabset}

```{r}
time_interest <- c("RPM1","RPM2", "RPM3", "RPM4")
#we plot here log10(value + 0.1)
expression_tumors <- function(object, gene){
  plot_interest <-  plot_cell_trajectory(object, markers = gene,  use_color_gradient = T) + facet_wrap(~time, nrow = 1)
plot_interest$data <- plot_interest$data %>% filter(time %in% time_interest)
subset_df <- plot_interest$data %>% filter(value>0)
  subset_df$new_val <- log10(subset_df$value + 0.1)
    pdf(file = paste0("/SSD/maa7095/scRNAseq/Resin_Esther_sc_small_cell_lung_cancer/SCLC_sc_analysis/Ireland_analysis/plots_pdf/new_script/Expr_tumors/",gene, ".pdf") ,  # The directory you want to save the file in
    width = 10, # The width of the plot in inches
    height = 10) # The height of the plot in inches

  plot2 <- plot_interest + 
    geom_point(data = subset_df, aes(x = data_dim_1, y = data_dim_2, colour = new_val)) + 
    scale_colour_gradient(low = "black", high = "green") + 
    labs(color = "Expression") + 
    theme(legend.key.size = unit(1, "cm"))
  print(plot2)
  dev.off()
}
```

```{r}
for (gene in genes_to_plot) {
  expression_tumors(pseudotime_object, gene)
}
```

# {-}

# NE profile in pseudotime coordinates for time points RPM1, RPM2, RPM3 and RPM4 with 0s removed. {.tabset}

```{r}
time_interest <- c("RPM1","RPM2", "RPM3", "RPM4")
#we plot here log10(value + 0.1)
NE_pseudotime_gene_tumors <- function(object, gene){
  plot_interest <-  plot_cell_trajectory(object, markers = gene,  use_color_gradient = F, color_by = "NE_spearman") + facet_wrap(~time, nrow = 1)
  plot_interest$data <- plot_interest$data %>% filter(time %in% time_interest)
  plot_interest$data <- plot_interest$data %>% filter(value > 0)
  plot_interest$data <-  plot_interest$data %>% drop_na(NE_spearman)
  plot_interest$mapping$size <- NULL
  
    pdf(file = paste0("/SSD/maa7095/scRNAseq/Resin_Esther_sc_small_cell_lung_cancer/SCLC_sc_analysis/Ireland_analysis/plots_pdf/new_script/NE_tumors_0s_removed/",gene, ".pdf") ,  # The directory you want to save the file in
    width = 10, # The width of the plot in inches
    height = 10) # The height of the plot in inches

  plot2 <- plot_interest +  scale_colour_gradient(low = "black", high = "brown3") +labs(color = "NE score")  +ggtitle(paste0(gene))+ theme(legend.key.size = unit(1, "cm"))
  
  print(plot2)
  dev.off()
}
```


```{r}
for (gene in genes_to_plot) {
  NE_pseudotime_gene_tumors(pseudotime_object, gene)
}
```

## {-}

## NE profile for each RPM.
```{r}
time_interest <- c("RPM1","RPM2", "RPM3", "RPM4")
#we plot here log10(value + 0.1)
  plot_interest_RPM <-  plot_cell_trajectory(pseudotime_object, use_color_gradient = F, color_by = "NE_spearman") + facet_wrap(~time, nrow = 1)
  plot_interest_RPM$data <- plot_interest_RPM$data %>% filter(time %in% time_interest)
  plot_interest_RPM$data <-  plot_interest_RPM$data %>% drop_na(NE_spearman)
    pdf(file="/SSD/maa7095/scRNAseq/Resin_Esther_sc_small_cell_lung_cancer/SCLC_sc_analysis/Ireland_analysis/plots_pdf/new_script/NE_tumors.pdf" ,  # The directory you want to save the file in
    width = 10, # The width of the plot in inches
    height = 10) # The height of the plot in inches

  plot2 <- plot_interest_RPM +  scale_colour_gradient(low = "black", high = "brown3") +labs(color = "NE score") + theme(legend.key.size = unit(1, "cm"))
  
  print(plot2)
  dev.off()

```



# {-}


Dimensionlaity reduction and clustering for RPM tumor cells.

```{r}
#create seurat object from pseudotime.
Ireland_mat_pseudo <- as.matrix(pseudotime_object@assayData$exprs)
Ireland_metadata_pseudo <- pseudotime_object@phenoData@data
```

```{r}
#subset just for RPM.
rpm_cells <- Ireland_metadata_pseudo %>% filter(time %in% c("RPM1", "RPM2", "RPM3", "RPM4")) %>% rownames()
```



```{r}
Ireland_seurat_pseudo <- CreateSeuratObject(Ireland_mat_pseudo, meta.data = Ireland_metadata_pseudo, assay = "RNA")
Ireland_seurat_pseudo[["RNA"]]$data <- Ireland_mat_pseudo
Ireland_seurat_pseudo[["RNA"]]$counts <- NULL

RPM_Ireland <- Ireland_seurat_pseudo[,rpm_cells]
```

```{r}
all.genes <- rownames(RPM_Ireland)
RPM_Ireland <- ScaleData(RPM_Ireland, features = all.genes)
RPM_Ireland <- RunPCA(RPM_Ireland, features = all.genes)
```


```{r}
RPM_Ireland <- FindNeighbors(RPM_Ireland, dims = 1:20)
RPM_Ireland <- FindClusters(RPM_Ireland, resolution = 0.5)
RPM_Ireland <- RunTSNE(RPM_Ireland, dims = 1:20)
```

```{r}
saveRDS(RPM_Ireland, "seurat_RPM_Ireland.rds")
```

```{r}
#;load R
RPM_Ireland <- readRDS("seurat_RPM_Ireland.rds")
```



# Two dimensionality plots. {.tabset}

## tSNE clustering for tumor RPM samples.
```{r}
pdf(file ="/SSD/maa7095/scRNAseq/Resin_Esther_sc_small_cell_lung_cancer/SCLC_sc_analysis/Ireland_analysis/plots_pdf/new_script/RPM_tsNE.pdf",   # The directory you want to save the file in
    width = 10, # The width of the plot in inches
    height = 10) # The he
dimplot <- DimPlot(RPM_Ireland, reduction = "tsne", group.by = "time")  + theme(
  axis.title.y = element_text(size = rel(1.5)),
  axis.title.x = element_text(size = rel(1.5)),
  axis.text.y = element_text(size = 20),
  axis.text.x = element_text(size = 20)) + ggtitle("")

print(dimplot)

dev.off()
```

```{r}
tsne_ireland <- as.data.frame(RPM_Ireland@reductions$tsne@cell.embeddings)
tsne_ireland$real_Barcode <- rownames(tsne_ireland)
smrca4_expr <- as.data.frame(RPM_Ireland["Smarca4",]@assays$RNA@layers$data)
names(smrca4_expr) <- "Norm_exp"
smrca4_expr$real_Barcode <- colnames(RPM_Ireland)
smrca4_expr <- join(smrca4_expr,tsne_ireland, type="inner")
#to color we are doing log10(value + 0.1) as seen in the monocle package
smrca4_expr$SMARCA4 <- log10(smrca4_expr$Norm_exp + 0.1)
```

## log10 SMARCA4 expression values.
```{r}
pdf(file ="/SSD/maa7095/scRNAseq/Resin_Esther_sc_small_cell_lung_cancer/SCLC_sc_analysis/Ireland_analysis/plots_pdf/new_script/tSNE_LogExp_Smarca4_RPM.pdf",   # The directory you want to save the file in
    width = 10, # The width of the plot in inches
    height = 10) # The he
 p1 <- ggplot(smrca4_expr) + geom_point(aes(x=tSNE_1, y=tSNE_2, color=SMARCA4)) 
  subset_df_smrca4_expr <- smrca4_expr %>% dplyr::filter(Norm_exp  > 0)
  p2 <- p1 + 
    geom_point(data = subset_df_smrca4_expr, aes(x = tSNE_1, y = tSNE_2, colour = SMARCA4)) +   scale_colour_gradient(low = "black", high = "green") + theme(
  axis.title.y = element_text(size = rel(1.5)),
  axis.title.x = element_text(size = rel(1.5)),
  axis.text.y = element_text(size = 20),
  axis.text.x = element_text(size = 20), 
  axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank())
  print(p2)
  dev.off()
```

```{r}
metadata_rpm <- RPM_Ireland@meta.data
metadata_rpm$real_Barcode <- rownames(metadata_rpm)
metadata_rpm <- plyr::join(metadata_rpm,tsne_ireland, type="inner" )
```

## NE score across the cells.
```{r}

metadata_rpm <- metadata_rpm %>% drop_na(NE_spearman)
pdf(file ="/SSD/maa7095/scRNAseq/Resin_Esther_sc_small_cell_lung_cancer/SCLC_sc_analysis/Ireland_analysis/plots_pdf/new_script/tSNE_RPM_NE.pdf",   # The directory you want to save the file in
    width = 10, # The width of the plot in inches
    height = 10) # The he
NE_plot <- ggplot(metadata_rpm) + geom_point(aes(x=tSNE_1, y=tSNE_2, color=NE_spearman)) +  scale_colour_gradient(low = "black", high = "brown3") + theme(
  axis.title.y = element_text(size = rel(1.5)),
  axis.title.x = element_text(size = rel(1.5)),
  axis.text.y = element_text(size = 20),
  axis.text.x = element_text(size = 20), 
  axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()) +labs(color = "NE score") 
print(NE_plot)
dev.off()
```

# {-}

Save sSmarca4 values and NE_spearman values.
```{r}
sub_metadata_rpm <- metadata_rpm %>% select(real_Barcode, NE_spearman, time)
big_smarca_table <- join(smrca4_expr, sub_metadata_rpm, type = "inner")
sub_smarca4 <- big_smarca_table %>% select(Norm_exp,tSNE_1, tSNE_2,  SMARCA4, NE_spearman, time)
write.csv(sub_smarca4,"Ireland_analysis/smarca4_NE_RPMs_values.csv", quote = F )
```

